\texttt{ensmallen}: a generic C++ library for fast optimization

### 2018-10-14 doc/mlsys_paper/paper.tex
Add references, slight refactoring.

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data. %TODO: cite
Even popular machine learning models such as logistic regression % TODO: cite
have training times mostly dominated by an optimization.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%% CS: i think we don't need to explore this too much; better to cut out all the
%% CS: fat and stick with concrete examples, instead of veering off on tangents
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?
%% CS: NIPS is too self-referential here

@@

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe, % TODO: cite
PyTorch, % TODO: cite
and TensorFlow % TODO: cite
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as {\tt scikit-learn} % TODO: cite
% TODO: can we find another library to put here too?  Shark maybe?
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, {\tt scipy.optimize} % TODO: check and cite
is widely used in the Python community, and MATLAB's {\tt TODO()}
% TODO: what's the name of that MATLAB function? or one of them?
% TODO: cite
has been available and used for many decades.  However, these implementations
are often not suitable for modern machine learning tasks---for instance,
computing the full gradient of the objective function may not be feasible
because there are too many data points.

@@

{\tt ensmallen} is open-source software licensed under the 3-claused BSD license~\cite{TODO},
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{TODO} is used for efficient linear algebra operations.

% Originally, {\tt ensmallen} was developed as part of {\tt mlpack}; % TODO: cite
% it is now available in standalone header-only form.
% TODO: seems awkward to put the lineage here;
%% CS: agreed; i don't see a need to explicitly mention mlpack -- it distracts
%% CS: from the central message
% TODO: also, we should talk up the use
% of Armadillo as its only dependency, which is nice, and maybe emphasize the
% utility of a header-only library

@@

We can implement these two functions in a class {\tt LinearRegressionFunction}.
Here {\tt arma::mat} and {\tt arma::vec} represent Armadillo matrix and vector
types, respectively. % TODO: cite Armadillo

@@

In essence, given a {\tt FunctionType} class that implements either an {\tt
Evaluate()} and {\tt Gradient()} method or an {\tt EvaluateWithGradient()}
method, template metaprogramming techniques are used to detect which of these
methods exist.  Then, a wrapper class will use suitable mix-ins % TODO: cite
in order to provide the `missing' functionality.  If {\tt
EvaluateWithGradient()} does not exist, the following code is generated:

@@

First, we consider the simple and popular Rosenbrock function: % TODO: cite

@@

To demonstrate the added overhead of other frameworks, we choose to use
simulated annealing, % TODO: cite
a gradient-free optimizer that only uses the objective function itself.
Simulated annealing will call the objective function very many times; for each
simulation we limit the optimizer to $100,000$ objective evaluations.  We
compare four frameworks for this task:

\begin{itemize}
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1 % TODO: cite
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia 1.0.1
% TODO: cite
  \item {\tt samin} from GNU Octave's {\tt optim} package % TODO: cite
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}

@@

Next, we consider the same machine learning example we have been working with
through the paper: linear regression.  For this task we will use the first-order
L-BFGS optimizer, % TODO: cite
so for {\tt ensmallen} we will implement two versions: one with {\tt Evaluate()}
and {\tt Gradient()}, and one with only {\tt EvaluateWithGradient()}.  The code
for each of these is the same as shown earlier.  In addition, we have more
options for Julia: we can implement only the objective function and allow either
% TODO: cite these two
Julia's {\tt Calculus.jl} or {\tt ForwardDiff.jl} package to automatically
compute the gradient.  Therefore, we compare against the following:

>>>>

Optimization is often computationally intensive and may correspond
to most of the time taken to train a machine learning model.  For instance, the
training of deep neural networks is dominated by the optimization of the model
parameters on the data~\cite{schmidhuber2015deep}.
Even popular machine learning models such as logistic regression
have training times mostly dominated by an optimization~\cite{kingma2015adam}.
% TODO: might be nice to have something kind of anecdotal like 'even new
% students to the field of machine learning quickly encounter optimization' and
% cite, e.g., Andrew Ng's coursera course or some ML textbook or similar
%% CS: i think we don't need to explore this too much; better to cut out all the
%% CS: fat and stick with concrete examples, instead of veering off on tangents
%
% or maybe just a note about how many optimization techniques get published at
% NIPS every year?
%% CS: NIPS is too self-referential here

@@

Although there are many existing machine learning optimization toolkits, few
are able to take explicit advantage of metaprogramming based code optimizations,
and few offer robust support for various types of objective functions.
For instance, deep learning
libraries like Caffe~\cite{jia2014caffe},
PyTorch~\cite{paszke2017automatic},
and TensorFlow~\cite{abadi2016tensorflow}
each contain a variety of optimization techniques.  However, these techniques are
limited to stochastic gradient descent (SGD) and SGD-like optimizers that
operate on small batches of data points at a time.  Other machine learning
libraries, such as {\tt scikit-learn}~\cite{pedregosa2011scikit}
% TODO: can we find another library to put here too?  Shark maybe?
contain optimization algorithms but not in a coherent or reusable framework.
Many programming languages have higher-level packages for
mathematical optimization.  For example, {\tt scipy.optimize} % TODO: check and cite
is widely used in the Python community, and MATLAB's function optimization
support has been available and used for many decades.  However, these
implementations are often unsuitable for modern machine learning tasks---for
instance, computing the full gradient of the objective function may not be
feasible because there are too many data points.

@@

{\tt ensmallen} is open-source software licensed under the 3-clause BSD
license\footnote{\url{https://opensource.org/licenses/BSD-3-Clause}},
allowing unencumbered use in both open-source and proprietary projects.
It is available for download from \url{https://www.ensmallen.org}.
Armadillo~\cite{sanderson2016armadillo} is used for efficient linear algebra operations.

% Originally, {\tt ensmallen} was developed as part of {\tt mlpack}; % TODO: cite
% it is now available in standalone header-only form.
% TODO: seems awkward to put the lineage here;
%% CS: agreed; i don't see a need to explicitly mention mlpack -- it distracts
%% CS: from the central message
% TODO: also, we should talk up the use

@@

We can implement these two functions in a class {\tt LinearRegressionFunction}.
Here {\tt arma::mat} and {\tt arma::vec} represent Armadillo matrix and vector
types, respectively~\cite{sanderson2016armadillo}.

@@

In essence, given a {\tt FunctionType} class that implements either an {\tt
Evaluate()} and {\tt Gradient()} method or an {\tt EvaluateWithGradient()}
method, template metaprogramming techniques are used to detect which of these
methods exist.  Then, a wrapper class will use suitable mix-ins
in order to provide the `missing' functionality~\cite{smaragdakis2000mixin}.  If
{\tt EvaluateWithGradient()} does not exist, the following code is generated:

@@

First, we consider the simple and popular Rosenbrock
function~\cite{Rosenbrock1960}:

@@

To demonstrate the added overhead of other frameworks, we choose to use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer that only uses the objective function itself.
Simulated annealing will call the objective function very many times; for each
simulation we limit the optimizer to $100,000$ objective evaluations.  We
compare four frameworks for this task:

\begin{itemize}
  \item {\tt ensmallen}
  \item {\tt scipy.optimize.anneal}, from scipy 0.14.1~\cite{jones2014scipy}
  \item {\tt Optim.jl}'s simulated annealing implementation with Julia
1.0.1~\cite{mogensen2018optim}
  \item {\tt samin} from GNU Octave's {\tt optim} package~\cite{octave}
\footnote{Another option here might be {\tt simulannealbnd()} from MATLAB's
Global Optimization Toolkit.  However, no license was available for these
simulations.}
\end{itemize}

@@

Next, we consider the same machine learning example we have been working with
through the paper: linear regression.  For this task we will use the first-order
L-BFGS optimizer~\cite{zhu1997algorithm},
so for {\tt ensmallen} we will implement two versions: one with {\tt Evaluate()}
and {\tt Gradient()}, and one with only {\tt EvaluateWithGradient()}.  The code
for each of these is the same as shown earlier.  In addition, we have more
options for Julia: we can implement only the objective function and allow either
Julia's {\tt
Calculus.jl}\footnote{\url{https://github.com/JuliaMath/Calculus.jl}} or {\tt
ForwardDiff.jl}~\cite{RevelsLubinPapamarkou2016} package to automatically
compute the gradient.  Therefore, we compare against the following:
